Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-06-12 newpaper/paper.tex
paper revision

Why not just make the object language less liberal, e.g.\ by disallowing
$\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
easier or more obvious? There is a trade-off between making the object language
more restricted, and thus easier to compile, and making \emph{metaprogramming}
more convenient. We will see that the ability to insert $\mbf{let}$-s without
restriction is very convenient in code generation, and likewise the ability to
have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
with the most liberal object syntax, at the cost of needing more downstream
processing. On the other side of the spectrum, one might imagine generating
object code in low-level A-normal form. This is more laborious, but it could be
also interesting, because it would force us to invent abstractions for
manipulating ANF in the metalanguage. We leave such lower-level object languages
for future investigation.

@@

In the meta level, we can borrow essentially all definitions from
\texttt{mtl}. Only the continuation monad transformer fails to support
binding-time-improvement in CFTT, because of the obvious need for dynamic
closures. In the following we only present $\StateT$ and $\ReaderT$.

@@

There is a deficiency in our library so far. Consider:
\begin{alignat*}{3}
  & \vf : \Bool \to \StateT\,\Nat\,(\MaybeT\,\Identity)\,()\\
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
This is unstaged to the following (omitting newtype wrappers):
\begin{alignat*}{3}
  & \vf\,\vb\,\vs := \case\,\vb\,\of \\
  & \ind \True\hspace{0.2em} \to \letdef \vs := 10; \letdef \vs := \vs + \vs; \Just\,((),\vs)\\
  & \ind \False \to \letdef \vs := 20; \letdef \vs := \vs + \vs; \Just\,((),\vs)
\end{alignat*}
Notice that the final state modification gets inlined into both branches. The
duplication follows from the definition of monadic binding in $\Gen$ and the
$\mit{split}$ function in the desugaring of $\mbf{case}$. Code generation is
continued in both branches with the same action. If we have multiple Boolean
$\mbf{case}$ splits sequenced after each other, that yields exponential code
size. A possible fix is to let-bind the branching action:
\begin{alignat*}{3}
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \msf{act} \fro \gen \dlr \mdown \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20} \\
  & \ind \mup\,\msf{act} \\
  & \ind\modify\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
However, this yields suboptimal code:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{act}\,\vs := \case\,\vb\,\of \True\hspace{0.2em}\to \letdef \vs := 10; \Just((), \vs)\\
  & \ind\hspace{8.5em}                               \False \to \letdef \vs := 20; \Just((), \vs);\\
  & \ind \case \msf{act}\,\vs\,\of\\
  & \ind\ind \Just\,(\_,\,\vs) \to \letdef \vs := \vs + \vs; \Just\,((),\,\vs)\\
  & \ind\ind \Nothing\hspace{0.6em} \to \Nothing
\end{alignat*}
This solution is fairly good in a state monad, where $\mdown$ only introduces a
runtime pair constructor, and it is feasible to compile object-level pairs as
unboxed data, without overheads. However, for $\Maybe$, $\mdown$ introduces a
runtime $\Just$ or $\Nothing$, and $\mup$ introduces a runtime case split. A
better solution would be to introduce two let-bound join points before
the offending $\mbf{case}$, one for returning a $\Just$ and one for returning
$\Nothing$, but fusing away the actual runtime constructors:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{joinNothing}\,\vs := \Nothing;\\
  & \ind \letdef\,\msf{joinJust}\,\hspace{1.8em}\vs := \letdef \vs := \vs + \vs; \Just\,((),\,\vs);\\
  & \ind \case \msf{b}\,\of\\
  & \ind \ind \True\hspace{0.2em}  \to \letdef \vs := 10; \msf{joinJust}\,\vs\\
  & \ind \ind \False \to \letdef \vs := 20; \msf{joinJust}\,\vs
\end{alignat*}
Here, $\msf{joinNothing}$ happens to be dead code, but it is easy to clean that up in
downstream compilation. Such ``fused'' returns are possible whenever we have a
$\Gen\,\vA$ action at the bottom of the transformer stack, such that $\vA$ is
isomorphic to a meta-level finite sum of value types. Recall that $\Gen\,\vA$ is
defined as $\{\vR : \VTy\} \to (\vA \to \Up \vR) \to \Up \vR$. Here, if $\vA$ is
a finite sum, we can rearrange $\vA \to \Up \vR$ to a finite product of
functions.

We could proceed with finite sums, but we will need finite
\emph{sums-of-products} ($\SOP$ in short) later in Section in
\ref{sec:stream-fusion}, so we develop SOP-s. SOPs have been used in generic and
staged programming; see e.g.\ \cite{sop} and
\cite{DBLP:conf/haskell/PickeringLW20}. We view $\SOP$-s as a Tarski-style
universe consisting of a type of descriptions and a way of interpreting
descriptions into $\MTy$ (``$\El$'' for ``elements'' of the type).
\begin{alignat*}{3}
  &\USOP : \MTy                    &&\ind \El_\SOP : \USOP \to \MTy \\
  &\USOP = \List\,(\List\,\VTy)    &&\ind \El_\SOP\,\vA = ...
\end{alignat*}
A type description is a list of lists of value types. We decode this to a sum of
products of value types. For example, $\El_\SOP\,[[\Bool,\,\Bool],\,[]]$ (using
Haskell-like list notation) is isomorphic to $\Either\,(\Lift\Bool,\,\Lift\Bool)\,()$.
$\USOP$ is closed under
value types, finite product types and finite sum types. For instance, we have
$\Either_\SOP : \USOP \to \USOP \to \USOP$ together with $\Left_\SOP : \El_\SOP\,\vA
\to \El_\SOP\,(\Either_\SOP\,\vA\,\vB)$, $\Right_\SOP : \El_\SOP\,\vB \to
\El_\SOP\,(\Either_\SOP\,\vA\,\vB)$ and a case splitting operation. It is more
convenient to work with type formers in $\MTy$, and only convert to $\SOP$
representations when needed, so we define a class for the representable types:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\tyclass\,\IsSOP\,(\vA : \MTy)\,\where} \\
  & \hspace{-5em}\ind \Rep      &&: \USOP \\
  & \hspace{-5em}\ind \mit{rep} &&: \vA \simeq \El_\SOP\,\Rep
\end{alignat*}
Above, $\vA \simeq \Rep$ denotes a record containing a pair of back-and-forth
functions, together with proofs (as propositional equalities) that they are
inverses. We will overload $\mit{rep}$ as the forward conversion function with
type $\vA \to \El_\SOP\,\Rep$, and write $\mit{rep}^{-1}$ for its inverse. Now we define
an isomorphic presentation of $\El_\SOP\,\vA \to \Lift \vR$ as a product of
object-level functions:
\begin{alignat*}{3}
  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \MTy}\\
  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\Lift(\mit{foldr}\,(\to)\,\vR\,\vA),\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
\end{alignat*}
\begin{alignat*}{3}
  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR\\
  &\mindex    &&: \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to (\El_\SOP\,\vA \to \Lift \vR)
\end{alignat*}
We omit here the definitions of $\mtabulate$ and $\mindex$. We will also need to
let-bind all functions in a $\mit{Fun}_{\SOP\Lift}$:
\begin{alignat*}{3}
  & \mathrlap{\mit{genFun}_{\SOP\Lift} : \{\vA : \USOP\} \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR}\\
  & \mit{genFun}_{\SOP\Lift}\,\{\Nil\}        && ()             &&= \return\,()\\
  & \mit{genFun}_{\SOP\Lift}\,\{\Cons\,\_\,\vA\}&& (\vf,\,\mit{\vfs}) &&= ({,})\,{\fmap}\,\gen\,\vf\,{\ap}\,\mit{genFun}_{\SOP\Lift}\,\{\vA\}\,\vfs
\end{alignat*}
We introduce a class for monads that support control flow joining.
\begin{alignat*}{3}
  & \tyclass\,\Monad\,\vM \RA \MonadJoin\,\vM\,\where\\
  & \ind \join : \IsSOP\,\vA \RA \vM\,\vA \to \vM\,\vA
\end{alignat*}
The most interesting instance is the ``base case'' for $\Gen$:
\begin{alignat*}{3}
  &\instance \MonadJoin\,\Gen\,\where\\
  &\ind \join\,\vma = \Gen \dlr \lam \vk.\,\runGen \dlr \mdo\\
  &\ind \ind \mit{joinPoints} \fro \mit{genFun}_{\SOP\Lift}\,(\mtabulate\,(\vk \circ \mit{rep}^{-1}))\\
  &\ind \ind \va \fro \vma\\
  &\ind \ind \return \dlr \mindex\,\mit{joinPoints}\,(\mit{rep}\,\va)
\end{alignat*}
Here we first convert $\vk : \vA \to \Lift \vR$ to a product of join points and
let-bind each one of them. Then we generate code that returns to the appropriate
join point for each return value. Other $\MonadJoin$ instances are straightforward. In
$\StateT_\M$, we need the extra $\IsSOP\,\vS$ constraint because $\vS$ is returned
as a result value.
\begin{alignat*}{3}
  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\MaybeT_\M\,\vM)\,\where\\
  & \ind \join\,(\MaybeT_\M\,\vma) = \MaybeT_\M\,(\join\,\vma)\\
  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\ReaderT_\M\,\vR\,\vM)\,\where\\
  & \ind \join\,(\ReaderT_\M\,\vma) = \ReaderT_\M\,(\join \circ \vma)\\
  & \instance\,(\MonadJoin\,\vM,\,\IsSOP\,\vS) \RA \MonadJoin\,(\StateT_\M\,\vS\,\vM)\,\where\\
  & \ind \join\,(\StateT_\M\,\vma) = \StateT_\M\,(\join \circ \vma)
\end{alignat*}
Now, the following definition yields the previously seen code output with
$\msf{joinNothing}$ and $\msf{joinJust}$.
\begin{alignat*}{3}
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \join \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
It might make sense to also have a simple ``desugaring'' rule that inserts a
$\join$ whenever we have a object-level case split with more than one branch. At
worst, this generates some noise and dead code that is easy to remove by
conservative code optimization.

@@

A pull stream is a meta-level specification of a state machine:
\begin{alignat*}{3}
  & \data \Step\,\vS\,\vA = \Stop\,|\,\Skip\,\vS\,|\,\Yield\,\vA\,\vS\\
  & \data \Pull\,(\vA : \MTy) : \MTy\,\where\\
  & \ind \Pull : (\vS : \MTy) \to \IsSOP\,\vS \RA \Gen\,\vS \to (\vS \to \Gen\,(\Step\,\vS\,\vA)) \to \Pull\,\vA
\end{alignat*}
In the $\Pull$ constructor, $\vS$ is the type of the internal state which is
required to be a sum-of-products of value types by the $\IsSOP\,\vS$
constraint. Borrowing terminology from Coutts \cite{DBLP:phd/ethos/Coutts11}, we
call each product in the state a \emph{state shape}.

The next field with type $\Gen\,\vS$ is the initial state, while transitions
are represented by the $\vS \to \Gen\,(\Step\,\vS\,\vA)$ field. Possible transitions
are stopping ($\Stop$), transitioning to a new state while outputting a value
($\Yield$) and making a silent transition to a new state ($\Skip$).

@@

\subsubsection*{Higher-order abstract syntax and logical frameworks.}
Hofmann's work on the semantics of higher-order abstract syntax (HOAS) is an
important precursor to our work \cite{Hofmann:1999:SAH:788021.788940}. It
anticipates many of the later developments in logical frameworks and two-level
type theories. In particular, we build on Hofmann's presheaf model for our 2LTT
semantics, and his sketch of adequacy for HOAS corresponds to our definition of
unstaging and its soundness proof. In general, there is a close correspondence
between logical frameworks that represent object languages in HOAS style
(e.g.\ \cite{Harper93lf}) and two-level type theories, and in some cases the two
kinds of presentations are inter-derivable \cite[Section
  3.3]{DBLP:journals/corr/abs-2302-08837}.

Cocon is a logical framework that has a dependently typed meta level
(``computation'' level) and an object language that is configured by a signature
\cite{DBLP:journals/corr/abs-1901-03378}. Cocon, like 2LTTs, can be used as a
dependently typed metaprogramming language, but there are several
differences. First, Cocon supports and emphasizes intensional analysis
(i.e.\ the ability to do structural induction on object code), which is made
possible through a contextual modality \cite{nanevski2008contextual}. In
contrast, intensional analysis is missing from 2LTTs. However, in Cocon, we can
only metaprogram object-level terms by specifying their object-level contexts
(i.e.\ possible free variables), while in 2LTTs the object-level contexts are
implicit and are only computed during code generation. The latter style is
more convenient for staged compilation applications, where intensional analysis
is rarely used and explicit context tracking would add a significant amount of
bureaucracy.

@@

Coutts developed pull stream fusion in depth in
\cite{DBLP:phd/ethos/Coutts11}. We borrowed the basic design and the basic
stream combinator definitions from there. This account is also close to existing
stream implementations in Haskell. Coutts characterized fusibility in a
non-staged setting, in terms of ``good consumers'' and ``good producers''.  This
scheme does not cover $\concatMap$.

@@

Also, rebuilding known abstractions in staged programming is valuable because it
provides a \emph{semantic explanation} of how abstractions can be compiled, and
provides insights that could be reused even in non-staged settings. For
instance, in this paper we demonstrated that compiling monadic code can be done
by using \emph{the same monads} in the metalanguage, extended with code
generation as an effect.

@@

\appendices
\section{Staging in CFTT}\label{appendix:A}

@@

\subsection{W-types}
So far we have stayed close to \cite{staged2ltt}, with only cosmetic
changes. Supporting W-types, as a way to get inductive families
\cite{inductivefamilies}, is the biggest extension here. We extend Section 5 of
\cite{staged2ltt} here. In the syntax of CFTT, $\vW$-types are as follows. The
formal specification of syntactic $\vW{-}\msf{elim}$ is rather gnarly, because
of the usage of explicit substitutions, but we write it out here anyways.

@@

This is because both $\ev\,(\Id\,t\,u)\,\gamma$ and $\re\,(\Id\,t\,u)\,\gamma'$
are mere propositions --- the latter because of equality reflection in the
syntax.  In particular, if we have a witness of $\re\,(\Id\,t\,u)\,\gamma'$, we
also have $\re\,t\,\gamma' = \re\,u\,\gamma'$ by equality reflection. Thus, the
rest of the relational interpretation of $\Id$ is also trivial. Another way to
think about this is that soundness expresses a form of canonicity, but if we
have equality reflection then syntactic identities are always canonical.

>>>>

Why not just make the object language less liberal, e.g.\ by disallowing
$\lambda$ under $\mbf{case}$ or $\mbf{let}$, thereby making call saturation
easier or more obvious? There is a trade-off between making the object language
more restricted, and thus easier to compile, and making metaprogramming more
convenient. We will see that the ability to insert $\mbf{let}$-s without
restriction is very convenient in code generation, and likewise the ability to
have arbitrary object expressions in $\mbf{case}$ bodies. In this paper we go
with the most liberal object syntax, at the cost of needing more downstream
processing.

%% On the other side of the spectrum, one might imagine generating
%% object code in low-level A-normal form. This is more laborious, but it could be
%% also interesting, because it would force us to invent abstractions for
%% manipulating ANF in the metalanguage. We leave such lower-level object languages
%% for future investigation.

@@

On the meta level, we can borrow essentially all definitions from
\texttt{mtl}. Only the continuation monad transformer fails to support
binding-time-improvement in CFTT, because of the obvious need for dynamic
closures. In the following we only present $\StateT$ and $\ReaderT$.

@@

There is a deficiency in our library so far. Consider:
\begin{alignat*}{3}
  & \vf : \Bool \to \StateT\,\Nat\,(\MaybeT\,\Identity)\,()\\
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
This is unstaged to the following (omitting newtype wrappers):
\begin{alignat*}{3}
  & \vf\,\vb\,\vs := \case\,\vb\,\of \\
  & \ind \True\hspace{0.2em} \to \letdef \vs := 10; \letdef \vs := \vs + \vs; \Just\,((),\vs)\\
  & \ind \False \to \letdef \vs := 20; \letdef \vs := \vs + \vs; \Just\,((),\vs)
\end{alignat*}
Notice that the final state modification gets inlined into both branches. This
duplication follows from the definition of monadic binding in $\Gen$ and the
$\mit{split}$ function in the desugaring of $\mbf{case}$. Code generation is
continued in both branches with the same action. If we have multiple Boolean
$\mbf{case}$ splits sequenced after each other, that yields exponential code
size. A possible fix is to let-bind the branching action:
\begin{alignat*}{3}
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \msf{act} \fro \gen \dlr \mdown \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20} \\
  & \ind \mup\,\msf{act} \\
  & \ind\modify\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
However, this yields suboptimal code:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{act}\,\vs := \case\,\vb\,\of \True\hspace{0.2em}\to \letdef \vs := 10; \Just((), \vs)\\
  & \ind\hspace{8.5em}                               \False \to \letdef \vs := 20; \Just((), \vs);\\
  & \ind \case \msf{act}\,\vs\,\of\\
  & \ind\ind \Just\,(\_,\,\vs) \to \letdef \vs := \vs + \vs; \Just\,((),\,\vs)\\
  & \ind\ind \Nothing\hspace{0.6em} \to \Nothing
\end{alignat*}
This solution is fairly good when we only have $\State$ or $\Reader$ effects, where
$\mdown$ only introduces a runtime pair constructor, and it is feasible to
compile object-level pairs as unboxed data, without overheads. However, for
$\Maybe$, $\mdown$ introduces a runtime $\Just$ or $\Nothing$, and $\mup$
introduces a runtime case split. A better solution would be to introduce two
let-bound join points before the offending $\mbf{case}$, one for returning a
$\Just$ and one for returning $\Nothing$, but fusing away the actual runtime
constructors:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{joinNothing}\,\vs := \Nothing;\\
  & \ind \letdef\,\msf{joinJust}\,\hspace{1.8em}\vs := \letdef \vs := \vs + \vs; \Just\,((),\,\vs);\\
  & \ind \case \msf{b}\,\of\\
  & \ind \ind \True\hspace{0.2em}  \to \letdef \vs := 10; \msf{joinJust}\,\vs\\
  & \ind \ind \False \to \letdef \vs := 20; \msf{joinJust}\,\vs
\end{alignat*}
Here, $\msf{joinNothing}$ happens to be dead code, but it is easy to clean that up in
downstream compilation. Such ``fused'' returns are possible whenever we have a
$\Gen\,\vA$ action at the bottom of the transformer stack, such that $\vA$ is
isomorphic to a meta-level finite sum of value types. Recall that $\Gen\,\vA$ is
defined as $\{\vR : \VTy\} \to (\vA \to \Up \vR) \to \Up \vR$. Here, if $\vA$ is
a finite sum, we can rearrange $\vA \to \Up \vR$ to a finite product of
functions.

We could proceed with finite sums, but we will need finite
\emph{sums-of-products} ($\SOP$ in short) later in Section in
\ref{sec:stream-fusion}, so we develop SOP-s. SOPs have been used in generic and
staged programming; see e.g.\ \cite{sop} and
\cite{DBLP:conf/haskell/PickeringLW20}. We view $\SOP$-s as a Tarski-style
universe consisting of a type of descriptions and a way of interpreting
descriptions into $\MTy$ (``$\El$'' for ``elements'' of the type).
\begin{alignat*}{3}
  &\USOP : \MTy                    &&\ind \El_\SOP : \USOP \to \MTy \\
  &\USOP = \List\,(\List\,\VTy)    &&\ind \El_\SOP\,\vA = ...
\end{alignat*}
A type description is a list of lists of value types. We decode this to a sum of
products of value types. For example, $\El_\SOP\,[[\Bool,\,\Bool],\,[]]$ (using
Haskell-like list notation) is isomorphic to $\Either\,(\Lift\Bool,\,\Lift\Bool)\,()$.
$\USOP$ is closed under
value types, finite product types and finite sum types. For instance, we have
$\Either_\SOP : \USOP \to \USOP \to \USOP$ together with $\Left_\SOP : \El_\SOP\,\vA
\to \El_\SOP\,(\Either_\SOP\,\vA\,\vB)$, $\Right_\SOP : \El_\SOP\,\vB \to
\El_\SOP\,(\Either_\SOP\,\vA\,\vB)$ and a case splitting operation. It is more
convenient to work with type formers in $\MTy$, and only convert to $\SOP$
representations when needed, so we define a class for the representable types:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\tyclass\,\IsSOP\,(\vA : \MTy)\,\where} \\
  & \hspace{-5em}\ind \Rep      &&: \USOP \\
  & \hspace{-5em}\ind \mit{rep} &&: \vA \simeq \El_\SOP\,\Rep
\end{alignat*}
Above, $\vA \simeq \Rep$ denotes a record containing a pair of back-and-forth
functions, together with proofs (as propositional equalities) that they are
inverses. We will overload $\mit{rep}$ as the forward conversion function with
type $\vA \to \El_\SOP\,\Rep$, and write $\mit{rep}^{-1}$ for its inverse.

In 2LTT literature, a metatype $\vA$ is called \emph{cofibrant} if $\vA \to \vB$ is
isomorphic to an object type whenever $\vB$ is isomorphic to an object type
\cite{twolevel}. We conjecture that our $\IsSOP$ types are exactly the cofibrant
types in this sense. However, ``cofibrant'' is not very descriptive in our
setting, so we shall keep writing $\IsSOP$.

Accordingly, we define an isomorphic presentation of $\El_\SOP\,\vA \to
\Lift \vR$ as a product of object-level functions:
\begin{alignat*}{3}
  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \MTy}\\
  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\Lift(\mit{foldr}\,(\to)\,\vR\,\vA),\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
\end{alignat*}
\begin{alignat*}{3}
  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR\\
  &\mindex    &&: \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to (\El_\SOP\,\vA \to \Lift \vR)
\end{alignat*}
We omit here the definitions of $\mtabulate$ and $\mindex$. We will also need to
let-bind all functions in a $\mit{Fun}_{\SOP\Lift}$:
\begin{alignat*}{3}
  & \mathrlap{\mit{genFun}_{\SOP\Lift} : \{\vA : \USOP\} \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR}\\
  & \mit{genFun}_{\SOP\Lift}\,\{\Nil\}        && ()             &&= \return\,()\\
  & \mit{genFun}_{\SOP\Lift}\,\{\Cons\,\_\,\vA\}&& (\vf,\,\mit{\vfs}) &&= ({,})\,{\fmap}\,\gen\,\vf\,{\ap}\,\mit{genFun}_{\SOP\Lift}\,\{\vA\}\,\vfs
\end{alignat*}
We introduce a class for monads that support control flow joining.
\begin{alignat*}{3}
  & \tyclass\,\Monad\,\vM \RA \MonadJoin\,\vM\,\where\\
  & \ind \join : \IsSOP\,\vA \RA \vM\,\vA \to \vM\,\vA
\end{alignat*}
The most interesting instance is the ``base case'' for $\Gen$:
\begin{alignat*}{3}
  &\instance \MonadJoin\,\Gen\,\where\\
  &\ind \join\,\vma = \Gen \dlr \lam \vk.\,\runGen \dlr \mdo\\
  &\ind \ind \mit{joinPoints} \fro \mit{genFun}_{\SOP\Lift}\,(\mtabulate\,(\vk \circ \mit{rep}^{-1}))\\
  &\ind \ind \va \fro \vma\\
  &\ind \ind \return \dlr \mindex\,\mit{joinPoints}\,(\mit{rep}\,\va)
\end{alignat*}
Here we first convert $\vk : \vA \to \Lift \vR$ to a product of join points and
let-bind each one of them. Then we generate code that returns to the appropriate
join point for each return value. Other $\MonadJoin$ instances are straightforward. In
$\StateT_\M$, we need the extra $\IsSOP\,\vS$ constraint because $\vS$ is returned
as a result value.
\begin{alignat*}{3}
  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\MaybeT_\M\,\vM)\,\where\\
  & \ind \join\,(\MaybeT_\M\,\vma) = \MaybeT_\M\,(\join\,\vma)\\
  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\ReaderT_\M\,\vR\,\vM)\,\where\\
  & \ind \join\,(\ReaderT_\M\,\vma) = \ReaderT_\M\,(\join \circ \vma)\\
  & \instance\,(\MonadJoin\,\vM,\,\IsSOP\,\vS) \RA \MonadJoin\,(\StateT_\M\,\vS\,\vM)\,\where\\
  & \ind \join\,(\StateT_\M\,\vma) = \StateT_\M\,(\join \circ \vma)
\end{alignat*}
Now, the following definition yields the previously seen code output with
$\msf{joinNothing}$ and $\msf{joinJust}$.
\begin{alignat*}{3}
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \join \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
It might make sense to also have a simple ``desugaring'' rule that inserts a
$\join$ whenever we have a object-level case split with more than one branch. At
worst, this generates some noise and dead code that is easy to remove by
conservative code optimization.

@@

A pull stream is a meta-level specification of a state machine:
\begin{alignat*}{3}
  & \data \Step\,\vS\,\vA = \Stop\,|\,\Skip\,\vS\,|\,\Yield\,\vA\,\vS\\
  & \data \Pull\,(\vA : \MTy) : \MTy\,\where\\
  & \ind \Pull : (\vS : \MTy) \to \IsSOP\,\vS \RA \Gen\,\vS \to (\vS \to \Gen\,(\Step\,\vS\,\vA)) \to \Pull\,\vA
\end{alignat*}
In the $\Pull$ constructor, $\vS$ is the type of the internal state which is
required to be a sum-of-products of value types by the $\IsSOP\,\vS$
constraint. The next field with type $\Gen\,\vS$ is the initial state, while
transitions are represented by the $\vS \to \Gen\,(\Step\,\vS\,\vA)$
field. Possible transitions are stopping ($\Stop$), transitioning to a new state
while outputting a value ($\Yield$) and making a silent transition to a new
state ($\Skip$).

Our definition roughly follows the non-staged stream fusion setup of
Coutts \cite{DBLP:phd/ethos/Coutts11}; the difference is the addition of
$\IsSOP$ and $\Gen$ in our version. Also, borrowing terminology from Coutts, we
call each product in the state a \emph{state shape}.

@@

\subsubsection*{Higher-order abstract syntax and logical frameworks.}
Hofmann's work on the semantics of higher-order abstract syntax (HOAS) is an
important precursor to our work \cite{Hofmann:1999:SAH:788021.788940}. It
anticipates many of the later developments in logical frameworks and two-level
type theories. In particular, we build on Hofmann's presheaf model for our 2LTT
semantics, and his sketch of adequacy for HOAS corresponds to our definition of
unstaging and its soundness proof. In general, there is a close correspondence
between logical frameworks that represent object languages in HOAS style
(e.g.\ \cite{Harper93lf}) and two-level type theories, and in some cases the two
kinds of presentations are inter-derivable \cite[Section
  3.3]{DBLP:journals/corr/abs-2302-08837}. In fact, our Agda implementation
of CFTT also uses a HOAS embedding of the object theory.

Cocon is a logical framework that has a dependently typed meta level
(``computation'' level) and an object language that is configured by a signature
\cite{DBLP:journals/corr/abs-1901-03378}. Cocon, like 2LTTs, can be used as a
dependently typed metaprogramming language, but there are several
differences. First, Cocon supports and emphasizes intensional analysis
(i.e.\ the ability to do structural induction on object code), which is made
possible through a contextual modality \cite{nanevski2008contextual}. In
contrast, intensional analysis is missing from 2LTTs so far. However, in Cocon,
we can only manipulate object terms at the meta level by specifying their
concrete object-level contexts (i.e.\ possible free variables), while in 2LTTs
the object-level contexts are implicit and are only computed during code
generation. The latter style is more convenient for staged compilation
applications, where explicit context tracking would add a significant amount of
bureaucracy.

Nonetheless, it seems to be possible and useful to have a
system which simultaneously supports intensional analysis through contextual
types and the 2LTT-style implicit contexts.

@@

Coutts developed pull stream fusion in depth in
\cite{DBLP:phd/ethos/Coutts11}. We borrowed the basic design and the basic
stream combinator definitions from there. This account is also close to existing
stream implementations in Haskell. We find it interesting that our staged
definitions are very close to the non-staged versions there; essentially all of
the additional complexity is compartmentalized in our $\SOP$ module.  Coutts
characterized fusibility in a non-staged setting, in terms of ``good consumers''
and ``good producers'', but this scheme does not cover $\concatMap$.

@@

Also, rebuilding known abstractions in staged programming is valuable because it
provides a \emph{semantic explanation} of how abstractions can be compiled, and
provides insights that could be reused even in non-staged settings. For
instance, in this paper we demonstrated that compiling monadic code can be done
by using the same monads in the metalanguage, extended with code
generation as an effect.

@@

\appendices
\section{Staged Semantics for CFTT}\label{appendix:A}

@@

\subsection{W-types}
So far we have stayed close to \cite{staged2ltt}, with only cosmetic
changes. Supporting W-types, as a way to get inductive families
\cite{whynotw}, is the biggest extension here. We extend Section 5 of
\cite{staged2ltt} here. In the syntax of CFTT, $\vW$-types are as follows. The
formal specification of syntactic $\vW{-}\msf{elim}$ is rather gnarly, because
of the usage of explicit substitutions, but we write it out here anyways.

@@

This is because both $\ev\,(\Id\,t\,u)\,\gamma$ and $\re\,(\Id\,t\,u)\,\gamma'$
are mere propositions --- the latter because of equality reflection in the
syntax.  In particular, if we have a witness of $\re\,(\Id\,t\,u)\,\gamma'$, we
also have $\re\,t\,\gamma' = \re\,u\,\gamma'$ by equality reflection. Thus, the
rest of the relational interpretation of $\Id$ is also trivial. Another way to
think about this: soundness expresses a form of canonicity, but if we have
equality reflection then terms with identity types are always canonical.
